---
title: "Project-2 Rough Work"
author: "Hannes"
date: "2023-03-06"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

# Introduction
Shopping data coming from the e-commerce website itself, containing a revenue variable (response variable) indicating whether a purchase was made or not and several predictor variables)

# Data preparation
change variable names to same style. Change all variables given as characters to factors and in the end change some specific variables to factors where it just makes sense. Look at all the factor variables. Merge small groups together to one group "other".
```{r namechange}
library(dplyr)
library(tidyr)
library(ggplot2)
library(pROC)
library(forcats)
library(ggcorrplot)
load("2_online_shopping.RData")
data = Data # rename to "data"
rm(Data)
# first change the name to snake_case
data = data %>% rename(purchase = Revenue, administrative = Administrative,
                       administrative_duration = Administrative_Duration, 
                       informational = Informational, informational_duration = Informational_Duration,
                       product_related = ProductRelated, product_related_duration = ProductRelated_Duration,
                       bounce_rates = BounceRates, exit_rates = ExitRates,
                       page_values = PageValues, special_day = SpecialDay, month = Month,
                       operating_systems = OperatingSystems, browser = Browser, region = Region,
                       traffic_type = TrafficType, visitor_type = VisitorType, weekend = Weekend) 
#str(data) # to see which variables are factors and which are numeric
data = data %>% mutate_if(is.character, as.factor) # change character variables to factors
data = data %>% mutate_if(is.integer, as.numeric) # change integers variables to numeric
data = data %>% mutate_if(is.logical, as.factor) # change logical variables to factors
data = data %>% mutate_at(c("visitor_type", "region", "browser", "operating_systems",
                            "traffic_type"), as.factor) # individually change some variables to factors

# loop over the factor variables and merge categories with small number of observations together
for (i in 1:length(data)) {
  if(is.factor(data[, i])) {
    tt = table(data[, i])
    data[, i] = fct_collapse(data[, i], "other" = names(tt[tt < 40]))
  }
}
```


# Graphical Data Exploration

Look at histograms of data

```{r histograms}
data_num = data %>% mutate_if(is.factor, as.numeric) # data with all variables as numeric
data_long <- gather(data_num, key = "predictor", value = "value")

# create histogram plot with facets
ggplot(data_long, aes(x = value)) + 
  geom_histogram() + 
  facet_wrap(~ predictor, scales = "free") 

```

We see that most factor variables have levels with only very few observations. If we fit a model containing a categorocal variable with few observations for a level, we fit a parameter for this level and hence a single parameter for very few observations. This can lead to high variance in our predictor. Hence it before we start creating models, for each categorical variable we merch small categories (<40) together. Also

# First model 

Look at all the factor variables. Merge small groups together to one group "other".
Then compare model with all categories and model with merged categories.

```{r}
gm_all = glm(purchase ~ ., data=data, family="binomial") # model with merged categories
AUC_eval <- function(gmodel,Data){
  set.seed(517)
  Folds <- matrix(sample(1:dim(Data)[1]), ncol=5)
  AUC <- rep(0,5)
  for(k in 1:5){
    train <- Data[-Folds[,k],]
    test <- Data[Folds[,k],]
    my_gm <- glm(gmodel$formula, family="binomial", data=train)
    test_pred <- predict(my_gm, newdata = test, type="response")
    AUC[k] <- auc(test$purchase,test_pred)
  }
  return(mean(AUC))
}
print(AUC_eval(gm_all, Data = data))

gm_inter = glm(purchase ~ . + weekend:visitor_type, data = data, family="binomial")
gm_square = glm(purchase ~. + I(page_values^2), data = data, family = "binomial")
gm_exp = glm(purchase ~. + I(1 / (1 + exp(product_related_duration))), data = data, family="binomial")
logdata = data %>% mutate(administrative = sqrt(administrative))
gm_log = glm(purchase ~ .,  data = logdata, family = "binomial")
print(AUC_eval(gm_inter, Data = data))
print(AUC_eval(gm_square, data))
print(AUC_eval(gm_exp, data))
```


Plot residuals of model with merged categories
```{r dataexploration}
# do facetwrap
data %>%
  mutate(res = resid(gm_all)) %>%
  mutate_if(is.factor, as.numeric) %>% 
  pivot_longer(-res) %>% 
  ggplot(aes(y = res, x = value)) + 
  facet_wrap( ~ name, scales = "free") + 
  geom_point() + 
  geom_smooth()
```

Look at cooks distance to detect outliers (do it in the end when we have final model and compare AUC with outlier and without)
```{r}
cooks_distance = cooks.distance(gm_all)
#plot(cd)
cook_data <- data.frame(observation = 1:length(cooks_distance), cooks_distance = cooks_distance)
# Plot Cook's distance using ggplot
ggplot(cook_data, aes(x = observation, y = cooks_distance)) +
  geom_point() +
  geom_hline(yintercept = 4/(length(cooks_distance)-2), color = "red", linetype = "dashed") +
  labs(x = "Observation", y = "Cook's distance") +
  ggtitle("Cook's distance plot") + 
  geom_point(x = 7607, y = cooks_distance[7607], col = "red")
which.max(cooks_distance)
data[7607, ]
```


We can see that maybe we need to include a quadratic dependency of page_values.
We create diagnostic plots with a linear model

```{r linearmodel}
lm_all = lm(purchase ~ ., data=data)
# do facetwrap
data %>% mutate(res = resid(lm_all)) %>% mutate_if(is.factor, as.numeric) %>%
  pivot_longer(-res) %>% 
  ggplot(aes(y = res, x = value)) + facet_wrap( ~ name, scales = "free") + 
  geom_point() + geom_smooth()
```

We can also look at the correlation between all the variables.
```{r correlationplot}
corr <- round(cor(data_num), 1)
ggcorrplot(corr, ggtheme = ggplot2::theme_gray)
```   



