---
title: "Project-6"
author: "Hannes Gubler"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  bookdown::html_document2:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

# Introduction

The COVID-19 pandemic has had a huge impact on countries worldwide, including the United States. Therefore it is crucial to analyze the underlying patterns of the pandemic to be better prepared for similar scenarios in the future.

In this report, we will work with daily Covid deaths data across all states of the United States. The data is from USAFacts and can be downloaded on their [website](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map). We will analyze the data using Principal Component Analysis (PCA), which is a popular technique to examine data with a high number of dimensions (in our case the dimension of our data is the number of days in between 22.01.2020 and 29.04.2023, hence 1194 dimensions). After performing the PCA, we will interpret the first three principal components of our data as well as the low dimensional plot, which displays the data across the first two principal components in our case.

# Data Preparation

Our data contains the cumulative number of Covid deaths for each state (observations in rows). In this report, we will use the following state abbreviations.

* Alabama: AL
* Alaska: AK 
* Arizona: AZ
* Arkansas: AR 
* California: CA 
* Colorado: CO 
* Connecticut: CT 
* Delaware: DE 
* Florida: FL 
* Georgia: GA 
* Hawaii: HI
* Idaho: ID 	
* Illinois: IL 	
* Indiana: IN 
* Iowa: IA 	
* Kansas: KS 
* Kentucky: KY 	
* Louisiana: LA 
* Maine: ME 
* Maryland: MD 
* Massachusetts: MA 
* Michigan: MI 
* Minnesota: MN 	
* Mississippi: MS 
* Missouri: MO 	
* Montana: MT 
* Nebraska: NE 
* Nevada: NV 
* New Hampshire: NH 
* New Jersey: NJ 
* New Mexico: NM 
* New York: NY 
* North Carolina: NC 
* North Dakota: ND 
* Ohio: OH 
* Oklahoma: OK 	
* Oregon: OR 	
* Pennsylvania: PA 
* Rhode Island: RI 
* South Carolina: SC 	
* South Dakota: SD 	
* Tennessee: TN 
* Texas: TX 
* Utah: UT 
* Vermont: VT 
* Virginia: VA 	
* Washington: WA 
* West Virginia: WV 
* Wisconsin: WI 
* Wyoming: WY 	

In Figure (\@ref(fig:dailydeaths)) we plot the total number of Covid deaths in the US per day, which we get from summing up all state deaths and then differencing the cumulative data.

```{r dataprep}
library("readr")
library("tidyverse")
library("RColorBrewer")
library("ggpubr")
data = read_csv("data/covid_deaths_usafacts.csv")

# remove unneeded columns
data = data %>%
  dplyr::select(-StateFIPS, -countyFIPS, -"County Name")
# group data by state instead of by country
data = data %>%
  group_by(State) %>%
  summarize_all(sum)
```

```{r dailydeaths, fig.cap="Daily Covid deaths in the US from 22.01.2020 until 29.04.2023."}
# nationwide data (sum up all states)
nation_data = data %>%
  dplyr::select(-State)
nation_data = colSums(nation_data)
# create data such that we have deaths per day instead of cumulative deaths over time
diff_nation_data = c(nation_data[1], diff(nation_data))
# plot the diffed nationdata
plot_data = as.data.frame(diff_nation_data)
plot_data$time = 1:length(diff_nation_data)
ggplot(data = plot_data, aes(x = time, y = diff_nation_data)) + 
  geom_line() + 
  scale_x_continuous(breaks = c(1, 183, 368, 550, 735, 917, 1102), 
                     labels = c("22.01.2020", "22.07.2020", "22.01.2021",
                                "22.07.2021","22.01.2022", "22.07.2022", 
                                "22.01.2023")) + 
  labs(x = "time", y = "covid deaths")
```

We observe very non-smooth data and some negative values (indicating a negative number of Covid deaths per day which is certainly not possible). This phenomenon occurs because, when counting the Covid deaths over time, certain death cases from a particular day may be subtracted due to counters believing they had overcounted on previous days.

To tackle this issue we apply smoothing to our data using Nadarya-Watson kernel regression with a Gaussian kernel and bandwidth 15 (where the days are numerated from 1 to 1194, where 1194 is the total number of days in our data). We apply smoothing to each state separately.

In Figure (\@ref(fig:smoothing)), we replicate the same plot as in Figure (\@ref(fig:dailydeaths)) to review our smoothing results.

```{r smoothing, fig.cap="Smoothed daily Covid deaths in the US from 22.01.2020 until 29.04.2023."}
# smooth the data of every state separately using nadarya watson smoother 
y = as.matrix(data[,-1])
for (i in 1:nrow(y)) {
  smoothed_states = ksmooth(x = 1:ncol(y), 
        y = y[i, ],
        bandwidth = 15,
        kernel = "normal")
  y[i, ] = smoothed_states$y
}
# create dataframe of smoothed data
smoothed_data = as.data.frame(y)
smoothed_data$state = data$State
# create nationdata as before but use the smoothed state data
nation_data = colSums(y)
# create data such that we have deaths per day instead of cumulative deaths over time
diff_nation_data = c(nation_data[1], diff(nation_data))
# plotthe diffed smoothed nationdata
plot_data = as.data.frame(diff_nation_data)
plot_data$time = 1:length(diff_nation_data)
ggplot(data = plot_data, aes(x = time, y = diff_nation_data)) + 
  geom_line() + 
  scale_x_continuous(breaks = c(1, 183, 368, 550, 735, 917, 1102), 
                     labels = c("22.01.2020", "22.07.2020", "22.01.2021",
                                "22.07.2021","22.01.2022", "22.07.2022", 
                                "22.01.2023")) + 
  labs(x = "time", y = "covid deaths")
```
The smoothed data already looks better and we got rid of negative values. So from now on we work with the smoothed values for the Covid death numbers. 

A further preprocessing step, to have more comparable death cases between the states, is to divide all observations from a state by its total population on 01.07.2022, taken from [Wikipedia](https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_population). Therefore our data contains now the cumulative number of Covid deaths per capita for each state from 22.01.2020 until 29.04.2023, displayed in Figure (\@ref(fig:normalize)).

```{r normalize, fig.cap="Cumulative Covid deaths per capita by states."}
# populations by state in the same order as in our data
populations = c(733583, 5074296, 3045637, 7359197, 39029342, 5839926, 3626205,
                671803, 1018396, 22244823, 10912876, 1440196, 3200517, 1939033,
                12582032, 6833037, 2937150, 4512310, 4590241, 6981974, 6164660, 
                1385340, 10034113, 5717184 , 6177957, 2940057, 1122867, 
                10698973, 779261, 1967923, 1395231, 9261699, 2113344, 3177772 ,19677151, 
                11756058, 4019800, 4240137, 12972008, 1093734, 5282634, 909824, 
                7051339, 30029572, 3380800, 8683619, 647064, 7785786, 5892539,
                1775156, 581381)
# divide observations from every state by its population. Last column of dataframe 
# must be excluded because its state indicator.
per_capita = sweep(smoothed_data[, -ncol(smoothed_data)], 1, populations, FUN = "/")
smoothed_data[, -ncol(smoothed_data)] = per_capita
# plot cumulative deaths by state by reshaping data to long format
data_long = gather(smoothed_data, key = "Day", value = "Cases", -state)
data_long$Day = as.Date(data_long$Day)
# create colour palette that contains very distinct colour for better 
# readability of the plot
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',] 
col_palette = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, 
                            rownames(qual_col_pals)))
set.seed(2)
col_palette = sample(col_palette, 51)
ggplot(data_long, aes(x = Day, y = Cases, color = state)) +
  geom_line() +
  scale_color_manual(values = col_palette) +
  labs(y = "Cases per Capita", x = "time" )
```

## Registration

Before performing the PCA, we want to align the starting date of the first wave across the states to have a more comparable first wave. As the first Covid wave did not spread out at the same time in each state, we define the individual beginning of the first wave in each state to be the point where we have one death per one million inhabitants in that state. Furthermore, in our analysis we only focus on the first wave and therefore stop 75 days after the beginning of the first wave in each state. In Figure (\@ref(fig:registration)) we verify that we more or less capture the first Covid wave for each state.

```{r registration, fig.height=5, fig.width=10, fig.cap="Covid deaths per capita for all US states during the first 75 days after a state hits the one death per million inhabitants mark the first time. In the right plot the ten states with the highest numbers during the above described period where excluded to better analyze the states with lower numbers."}
# normalize original data by population and store it in data_per_capita
# we use the original data to find the starting point of the first wave
per_capita = sweep(data[, -1], 1, populations, FUN = "/")
data_per_capita = data 
data_per_capita[, -1] = per_capita
# find the starting point for the first wave for each state and store 
# the next 75 observations from the starting point for each state in the 
# matrix first_wave_dat
start_ind = rep(0, 51)
n = 75
first_wave_dat = matrix(ncol = n, nrow = 51)
for (i in 1:51) {
  start_ind = max(which(data_per_capita[i, ] <= 10^-6))
  first_wave_dat[i, ] = as.numeric(smoothed_data[i, (start_ind + 1):(start_ind + n)])
}
# difference the first wave data again for a plot
diff_first_wave = apply(first_wave_dat, 1, diff)
diff_first_wave = as.data.frame(diff_first_wave)
# Add the x values for the plot
diff_first_wave$x = 1:(n-1)
# add state column names
colnames(diff_first_wave)[1:51] = data$State
# Reshape the dataframe to long format
long_data = gather(diff_first_wave, key = "state", value = "value", -x)
# repeat the steps for the same plot but without states that have 
# very high numbers, so we can analyze also the states with low numbers.
# start from second column because first column is the x-values
high_covid_states = names(sort(colSums(diff_first_wave), decreasing = TRUE))[2:11]
diff_first_wave_without_big = diff_first_wave[, !(names(diff_first_wave)) %in% 
                                                high_covid_states]
long_data_without_big = gather(diff_first_wave_without_big, 
                               key = "state", value = "value", -x)
# create colur palette for dataframe without the 10 most high covid states such that 
# the states from that plot have the same colurs as the states in the other plots
removed_state_ind = which(data$State %in% high_covid_states)
col_palette_removed = col_palette[-removed_state_ind]
# create the two plots
plot_all_states = ggplot(long_data, aes(x = x, y = value, color = state)) + 
  geom_line() +
  scale_color_manual(values = col_palette) + 
  labs(x = "days", y = "cases per capita")
plot_without_big = ggplot(long_data_without_big, 
                          aes(x = x, y = value, color = state)) + 
  geom_line() +
  scale_color_manual(values = col_palette_removed) +
  labs(x = "days", y = "cases per capita")
# plot the two plots next to each other
ggarrange(plot_all_states, plot_without_big, ncol = 2, nrow = 1, 
          common.legend = TRUE, legend = "right")
```

We observe that our time registration more or less captured the first wave for each state and therefore we proceed with this data.

In our PCA we will analyze the cumulative death numbers per capita from the data displayed in Figure (\@ref(fig:registration)) (Cumulative death numbers per capita during the first Covid wave). To have the data more spread out, we take the logarithm of the cumulative death numbers per capita as a last preprocessing step before we perform the PCA.

```{r log}
# apply log to the data
first_wave_dat = log(first_wave_dat)
```

# PCA

In this section we apply PCA to the data we prepared in the section before, that is the logarithm of the cumulative death numbers per capita during the first Covid wave for each state. In Figure (\@ref(fig:PCA)) we plot the first three principal components as well the percentages of variability retained.

```{r PCA, fig.width=10, fig.height=3.3, fig.cap="First three principal components of logged cumulative death numbers per capita during the first wave of Covid for each state as observations."}
derivative = apply(first_wave_dat, 1, function(row) diff(row))
derivative = t(as.matrix(diff_first_wave[, -ncol(diff_first_wave)]))
svd_mat = first_wave_dat
svd_mat_der = t(derivative)
# Center matrix by columns
mu = colMeans(svd_mat)
mu_sec_der = colMeans(svd_mat_der)
svd_mat_centered = scale(svd_mat, center = TRUE, scale = FALSE)
svd_mat_der_cent = scale(svd_mat_der, center = TRUE, scale = FALSE)
# calculate svd
svd = svd(svd_mat_centered)
svd_der = svd(svd_mat_der_cent)
# calculate variance explained
var_expl = svd$d^2 / (sum(svd$d^2))
var_expl_der = svd_der$d^2 / (sum(svd_der$d^2))
eig_val = as.data.frame(svd$v)
eig_val_der = as.data.frame(svd_der$v)
eig_val$x = 1:nrow(eig_val)
eig_val_der$x = 1:nrow(eig_val_der)
pc_1 = ggplot(data = eig_val, aes(x = x, y = V1)) + 
  geom_line() +
  labs(x = "days", y = "") + 
  ggtitle("1st PC (95.5% of var)")
pc_2 = ggplot(data = eig_val, aes(x = x, y = V2)) + 
  geom_line() +
  labs(x = "days", y = "") + 
  ggtitle("2nd PC (4.0% of var)")
pc_3 = ggplot(data = eig_val, aes(x = x, y = V3)) + 
  geom_line() +
  labs(x = "days", y = "") + 
  ggtitle("3rd PC (<1% of var)")
ggarrange(pc_1, pc_2, pc_3, nrow = 1, ncol = 3)
```

First note that the percentages of variability retained for the first component is very high (95.5%), so the first principal component already explains a big part of the variability in the data. Such a high number is possible since we only looked at a rather short time frame (75 days) instead of at the full data.

We can interpret the first three principal components as follows.

* The first principal component is all positive and can therefore be seen as the variability in overall cumulative Covid deaths in the first wave.
* The second principal component contrasts the beginning against the end of the first Covid wave.
* The third principal component contrasts the beginning and the end of the first wave with the middle (around the peak) of the first wave.

```{r PCAfirstder, include = FALSE}
pc_1_der = ggplot(data = eig_val_der, aes(x = x, y = V1)) + 
  geom_line() +
  labs(x = "days", y = "") + 
  ggtitle("1st PC (62% of var)")
pc_2_der = ggplot(data = eig_val_der, aes(x = x, y = V2)) + 
  geom_line() +
  labs(x = "days", y = "") + 
  ggtitle("2nd PC (17% of var)")
pc_3_der = ggplot(data = eig_val_der, aes(x = x, y = V3)) + 
  geom_line() +
  labs(x = "days", y = "") + 
  ggtitle("3rd PC (9% of var)")
ggarrange(pc_1_der, pc_2_der, pc_3_der, nrow = 1, ncol = 3)
```

## Low-Dimensional Plot

In PCA, the low-dimensional plot is the visualization of the data in a reduced dimensional space. In our case, the representation of the data as a linear combination of the first two principal components (which capture already more than 99% of the variability in the data). Furthermore we colorize the observations of our data (states) by the state's Democratic support in the US presidential election in 2020 (the data can be found on [Wikipedia](https://en.wikipedia.org/wiki/2020_United_States_presidential_election)). In Figure (\@ref(fig:lowdimplot)) we plot the low dimensional plot for our data colorized by Democratic support.

```{r lowdimplot, fig.cap="Low-dimensional plot of the first two principal components of the data containing the first Covid wave in the US, colorized by Democratic support in the 2020's US presidential eleciction."}
# calculate scores
scores = svd$u %*% diag(svd$d)
# store (rounded) voting results to colour them on plot
support_democratic <- c(
  0.43,   # AK
  0.37,   # AL
  0.35,   # AR
  0.49,   # AZ
  0.63,   # CA
  0.55,   # CO
  0.59,   # CT
  0.92,   # DC
  0.59,   # DE
  0.48,   # FL
  0.49,   # GA
  0.64,   # HI
  0.45,   # IA
  0.33,   # ID
  0.58,   # IL
  0.41,   # IN
  0.42,   # KS
  0.36,   # KY
  0.40,   # LA
  0.66,   # MA
  0.65,   # MD
  0.53,   # ME
  0.51,   # MI
  0.52,   # MN
  0.41,   # MO
  0.41,   # MS
  0.41,   # MT
  0.49,   # NC
  0.32,   # ND
  0.39,   # NE
  0.53,   # NH
  0.57,   # NJ
  0.54,   # NM
  0.50,   # NV
  0.61,   # NY
  0.45,   # OH
  0.32,   # OK
  0.56,   # OR
  0.50,   # PA
  0.59,   # RI
  0.43,   # SC
  0.36,   # SD
  0.37,   # TN
  0.46,   # TX
  0.38,   # UT
  0.54,   # VA
  0.66,   # VT
  0.58,   # WA
  0.49,   # WI
  0.30,   # WV
  0.27    # WY
)

# plot the results coloured by democratic support in 2020
plot_data = data.frame(x = scores[, 1], y = scores[, 2], color = support_democratic, states = data[, 1])
ggplot(plot_data, aes(x = x, y = y, color = support_democratic, label = State)) +
  geom_point() +
  geom_text(vjust = 1.5) +
  scale_color_gradient(low = "yellow", high = "darkblue") +
  labs(color = "Democratic Support Level", x = "1st PC (95.5% of var)", y = "2nd PC (4.0% of var)")

```
By colorizing the low-dimensional plot of our PCA we detect some grouping in the states that voted more Republic and states that voted more Democratic in the 2020 US presidential election. This suggests that there could be some connection between the development of the first Covid wave in a given state and its political ideology (Republic/Democratic). Especially the first principal component, which captures 95.5% of variability in the data, shows a clear pattern of Democratic states being more on the left side and  Republic states being more on the right side.

# Conclusion

In this report we analyzed data containing the cumulative number of deaths during Covid in the United States, and applied PCA to the first wave of the pandemic. To bring the first Covid wave, that behaved differently in each state, on the same scale we used registration by defining an individual starting point of the first wave in each state. After that, we kept the data containing the next 75 days after the registered starting point in each state to capture the first wave. After performing PCA, we found that our first two principal components captured almost all variability of the data, which makes it very interesting to look at the low-dimensional plot that represents the data in the first two principal components. In this plot, by colorizing the states with the Democratic support in the 2020 US presidential election, we found a connection between the first Covid wave in the US and the political state ideology. This is an interesting result, but as PCA is just an exploratory tool, a more in depth statistical analysis would be needed to draw definitive conclusions.